---
title: "Abundance of mountain yellow-legged frog across the Sierra Lakes"
author: "Quin Smith"
date: "1/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#read in packages
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(patchwork)
library(lubridate)
library(ggrepel)
```

### Introduction
This visual analysis with associated R-markdown file explores abundance data of mountain yellow-legged frogs (Rana muscosa) recorded by R.A. Knapp et al. and the Sierra Lakes Inventory Project.  The data was collected between 1995 and 2002 across a study area containing "over 8,000 lentic water bodies in the southern Sierra Nevada, including lakes, ponds, marshes, and meadows" (Environmental Data Initiative Repository).

```{r, echo = FALSE, message = FALSE,warning = FALSE}
#read in/wrangle data to only include mountain yellow-legged frogs (RAMU) across all water bodies by life stage and year, excluding 'EggMass'
frogs <- read_excel(here("data", "sierra_amphibians.xlsx"))

ramu_frogs <- frogs %>% 
  mutate(survey_date = ymd(survey_date)) %>% 
  mutate(year = year(survey_date)) %>% 
  filter(amphibian_species == "RAMU" & amphibian_life_stage != "EggMass") %>% 
  group_by(amphibian_life_stage, year) %>% 
  summarize(n = n())

#ggrepel label
gg_labeller <- ramu_frogs %>% 
  mutate(gg_lab = case_when(
    year == 2002 & amphibian_life_stage == "Adult" ~ "Adult",
    year == 2002 & amphibian_life_stage == "SubAdult" ~ "Sub-Adult",
    year == 2002 & amphibian_life_stage == "Tadpole" ~ "Tadpole"
  ))

#graph of number of individuals observed given conditions above
fig_palette <- c("#009E73", "#0072B2", "#D55E00")
fig_palette2 <- c("#CC6666", "#9999CC", "#66CC99")
 
ramu_graph <- ramu_frogs %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line(aes(color = amphibian_life_stage), size = 1) +
  geom_point(aes(color = amphibian_life_stage), size = 1.5) +
  scale_color_manual(values = fig_palette2) +
  labs(x = "Year", y = "Individuals observed", color = "Life Stage" ) +
  geom_label_repel(data = gg_labeller, aes(label = gg_lab),
                   nudge_x = -2,
                   na.rm = TRUE) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r, echo = FALSE, message = FALSE,warning = FALSE}
#wrangle data to create a subset of total counts of observed adult and subadult RAMU by lake, 5 top lakes (in observations)
ramu_lake <- frogs %>% 
  mutate(lake_id = paste("Lake", frogs$lake_id, sep = " ")) %>% 
  filter(amphibian_species == "RAMU" & 
           amphibian_life_stage %in% c("SubAdult", "Adult")) %>% 
  group_by(lake_id) %>% 
  summarize(n = n()) %>% 
  mutate(lake_id = fct_reorder(lake_id, n)) %>% 
  slice_max(order_by = n, n = 5)

#column graph with 5 top lakes
lake_graph <- ggplot(data = ramu_lake, aes(x = lake_id, y = n)) +
  geom_col(fill = "#CC6666") +
  labs(x = "Lake (ID for confidentiality)", y = "Individuals observed") +
  theme_bw()
```

```{r, echo = FALSE, message = FALSE,warning = FALSE, fig.width = 9, fig.height = 6, fig.cap = "**Figure 1:** a)  Total mountain yellow-legged frog (Rana muscosa) observations each year across all water bodies in the study area, 1995-2002.  b)  The 5 lakes with the greatest number of mountain yellow-legged frog observations (low to high).  Excludes tadpole observations.  Data: Sierra Lakes Inventory Project." }
#combine graphs
comp_graph <- ramu_graph + lake_graph +
  plot_annotation(tag_levels = 'a',
                  tag_suffix = ")")

comp_graph
```

### Data Citation:

Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3





















